<!--
%\VignetteEngine{knitr::knitr}
%\VignetteDepends{knitr}
%\VignetteIndexEntry{phyViz: Phylogenetic Visualization}

-->

How to use the package phyViz
========================================================

The very first thing to do is to load the necessary libraries and the example tree (tree.rda). 

```{r}
library(phyViz)
data(tree)
```

Now that you have loaded the raw tree.rda file, you can observe the first six lines to see the structure:

```{r}
head(tree)
```

Each line provides a relationship between a parent and a child; additional data fields (year, yield, etc.) generally refer to the child, with the exception of parent.type, which differentiates the male and female parents. 

You can convert it to the treeGraph format that will be used in subsequent functions:

```{r}
treeGraph <- processTreeGraph(tree)
head(treeGraph)
```

treeGraph format distills the tree dataset into essential components: the two nodes (parent, child) connected by an edge, and a weight for that edge (defaults to 1). 

It is generally useful to then convert the data frame object into an object that can be read by igraph package: 

```{r}
library(igraph)
mygraph <- graph.data.frame(treeGraph, directed= T)
```
